What We Eval in the Shadows

### 2021-04-15 paper/main.tex
ck's proofreading for Section 5.1 to 5.3

\medskip\noindent\EE{$min(e)=\c{X}$} Variables lookups are the most common form,
they are found in 28\% of the sites. This form includes simple variable reads,
\eg~\c{x}, those are \packageNbSymbolVarSitePercent of \c X. The form also
subsumes \c{V}, so it includes a mixture of arithmetic expressions,
\eg~\c{x+y+1}. The operations allowed are limited to built-in arithmetics. It is
noteworthy that, while most \c{X}\!s evaluate in a single step, the variable can
be bound to a promise, and accessing it may trigger evaluation of that promise,
thus resulting in an arbitrary amount of computation. The median number of
interpreted operations is \packageMinimizedmedianoperationsaRnd, suggesting that
it is not the common case. Lookups are often evaluated in constructed
environments, as few as \packageMinimizedpercentparentframesc of these
expressions are evaluated in a function environment.

@@

\medskip\noindent\EE{$min(e)=$~\c{<-}} This form includes both assignments
operator, the direct assignment {\tt <-}, and assignment to the parent
environment {\tt <\,\!<-}, and the \c{\$} form. Assignments occur in 4\% of the
sites. They represent the most obvious source of side-effects. The median number
of operations is \packageMinimizedmedianoperationsiRnd; the minimum is 3.

\medskip\noindent\EE{$min(e)=\c{F()}$} This form captures simple function calls,
\eg~\c{f(2)}, with neither variables or assignments. Or, more specifically it
allows for variables in the function position but not in arguments. Usually,
looking up function names does not trigger computation, but that is not a given
if the function is returned by a promise or if the function name is shadowed by
a promise containing a value, then computation will occur. This form occurs in
6\% of sites and typically does not perform much work in the interpreter.

\medskip\noindent\framebox{$min(e)=\c{F(X)}$}~\EE{$min(e)=\c{F(F(X))}$} These
forms allow for function calls whose arguments may include variable references.
The latter allows nested calls.Assignments are excluded from this form. They
occurs in, respectively, 14\% and 20\% of the sites in the corpus. Together they
are the most frequent forms. The median numbers of interpreter steps are,
respectively, \packageMinimizedmedianoperationsdRnd and
\packageMinimizedmedianoperationsbRnd. The \packageMinimizedpercentparentframesj
of these expressions run in function environments.

\medskip\noindent\EE{$min(e)=\c{FUN}$} This form captures expressions define
functions, \eg~\c{function(x)x+1}, and do nothing else. \c{FUN} occurs in
\packageFunctionDefinitionSitesPercent of sites. Evaluating a function
definition is done in 2 interpreter step; the data does not record the work
performed by the interpreter when the generated functions are eventually run. In
addition to \c{FUN}, \packageGeneralizedFunctionDefinitionSitesPercent of sites
have function definitions nested in other expressions, this gives an idea of the
use of higher-order functions.

\medskip\noindent\EE{$min(e)=\c{BLOCK}$} This form captures multi-statements
code blocks, which occur in only 4\% of sites. These are larger expressions, we
do not inspect contents of the blocks. The median number of executed operations
is \packageMinimizedmedianoperationsjRnd. They typically run in function
environments.

@@

The environment in which evaluation happens determines what is visible to the
computation started by \eval and the potential reach of its side-effects. The
second argument to \eval specifies that environment. Environment of \eval can be
classified into the following four kinds:

\begin{compactitem}[---]
\item \emph{Function:} environment for the local variables of some function
  currently active on the call stack. Obtained by calling \c{parent.frame()} or
  \c{sys.frame()}.
\item \emph{Synthetic:} environments built from data structures such as lists,
  dataframes, or constructed explicitly with \c{new.env}, \c{list2env} or
  \c{as.environment}. Also includes the empty environment.
\item \emph{Global:} the environment in which scripts or interactive commands
  are evaluated.
\item \emph{Package:} the environment of a loaded library.
\end{compactitem}

\noindent
As shown by Table~\ref{tab:highlevelenvironments}, most calls evaluate in a
function environment with the \c{global} kind as a distant second. This means
that most variable lookups and most side effects either read/update existing
local variables or introduce new ones. But for which function? From the point of
view of the function that called \eval, Table~\ref{tab:funoffset} gives an
offset on the call stack. Thus 0 is the direct caller and corresponds to local variables; 1 is its parent, and so on. The data suggests that in 81\% of cases, \eval access its caller's environment -- this means, the variables of the function where \eval textually
occurs are read and written to. It is interesting that some 1.5\% of sites
evaluate a code in an environment that is three frame or above from \eval. This
distance implies that, in general, modular reasoning is impossible, to
understand the behavior of any piece of code entails fully understanding the
behavior of all functions that the code may call as the actions of \eval may
happen at a distance. Finally note that any particular site may have several
kinds of environments, but in \packageNbOneCategoryEnvirSitePercent of the case a
site has a single kind.

@@

\paragraph{Discussion}
The data presented in this section is what one could expect. ({\it Function})
The majority of \eval call sites access the environment of some function on the
call stack. Usually, the current function, but sometimes frames arbitrarily far
up the call stack. One thing the data does not say is how that environment was
obtained. The expected case is that the environment was the one obtained from a
promise; as promises combine code and the environment in which that code
originates. Likely less frequent are cases where \eval is provided the results
of programmatically selecting some call stack. ({\it Synthetic}) The relative
high frequency of synthetic environments corresponds to cases when one wants to
evaluate an expression in either a restricted environment, use a data
structure as an environment. ({\it Global}) The fact that global environment
show up so frequently is likely an artifact of the way the code is run. ({\it
  Package}) There are only 92 sites that use a package environment in \eval,
this is probably best as it is bad form to add bindings or modify functions
of a loaded package. Many more sites rather wrap a package environment in a new
environment to create singletons in a package.

@@

\begin{compactitem}[---]
\item {\it Reflection:} This group corresponds to uses of the \c{match.call}
  function to reflectively capture the expression that invoked the current
  function.
\item {\it Constructed:} Expressions can be constructed by invoking the
  \c{quote}, \c{enquote}, \c{expression}. Function arguments are passed as
  promises and \c{substitute} is used to retrieve the source expression
  associated to the promise.
\item {\it String:} Finally, expressions can be created from strings by invoking
   \c{parse}, \c{str2expression} or \c{str2lang}.
\end{compactitem}

\noindent
Table~\ref{tab:provenance} summarizes the expression provenance in our corpus.
This data is obtained by dynamically tainting values as they are produced by the
various sources, due to technical reasons there is some imprecision in the
results. In particular, we are not able to classify all sites. Manual inspection
of numerous examples suggest that the classified results are accurate.

@@

\paragraph{Discussion}
The origin data suggests that constructing expressions from strings is a
minority of the use cases, instead the constructed category shows that the
majority of \evals comes code that was processed by the compiler, and may be
slightly modified by the programmer before invoking \eval. Both constructed and
reflection categories roughly correspond to meta-programming. Some of these use
cases could likely be replaced by macros if the R designers could be convinced
to overcome their distaste for those.

>>>>

\medskip\noindent\EE{$min(e)=\c{X}$} Variable lookups are the most common form,
they are found in 28\% of the sites. This form includes simple variable reads,
\eg~\c{x}, those are \packageNbSymbolVarSitePercent of \c X. The form also
subsumes \c{V}, so it includes a mixture of arithmetic expressions,
\eg~\c{x+y+1}. The operations allowed are limited to built-in arithmetics. It is
noteworthy that, while most \c{X}\!s evaluate in a single step, the variable can
be bound to a promise, and accessing it may trigger evaluation of that promise,
thus resulting in an arbitrary amount of computation. The median number of
interpreted operations is \packageMinimizedmedianoperationsaRnd, suggesting that
it is not the common case. Lookups are often evaluated in constructed
environments, as few as \packageMinimizedpercentparentframesc of these
expressions are evaluated in a function environment.

@@

\medskip\noindent\EE{$min(e)=$~\c{<-}} This form includes both assignment
operators, the direct assignment {\tt <-}, and assignment to the parent
environment {\tt <\,\!<-}, the \c{assign} function and the \c{\$} form. Assignments occur in 5\% of the
sites. They represent the most obvious source of side-effects. The median number
of operations is \packageMinimizedmedianoperationsiRnd; the minimum is 3.

\medskip\noindent\EE{$min(e)=\c{F()}$} This form captures simple function calls,
\eg~\c{f(2)}, with neither variables nor assignments. Or, more specifically it
allows for variables in the function position but not in arguments. Usually,
looking up function names does not trigger computation, but that is not a given
if the function is returned by a promise or if the function name is shadowed by
a promise containing a value, then computation will occur. This form occurs in
6\% of sites and typically does not perform much work in the interpreter.

\medskip\noindent\framebox{$min(e)=\c{F(X)}$}~\EE{$min(e)=\c{F(F(X))}$} These
forms allow for function calls whose arguments may include variable references.
The latter allows nested calls. Assignments are excluded from this form. They
occur in, respectively, 14\% and 20\% of the sites in the corpus. Together they
are the most frequent forms. The median numbers of interpreter steps are,
respectively, \packageMinimizedmedianoperationsdRnd and
\packageMinimizedmedianoperationsbRnd. The \packageMinimizedpercentparentframesj
of these expressions run in function environments.

\medskip\noindent\EE{$min(e)=\c{FUN}$} This form captures expressions that define functions, \eg~\c{function(x) x+1}, and do nothing else. \c{FUN} occurs in
\packageFunctionDefinitionSitesPercent of sites. Evaluating a function
definition is done in 2 interpreter step; the data does not record the work
performed by the interpreter when the generated functions are eventually run. In
addition to \c{FUN}, \packageGeneralizedFunctionDefinitionSitesPercent of sites
have function definitions nested in other expressions, this gives an idea of the
use of higher-order functions.

\medskip\noindent\EE{$min(e)=\c{BLOCK}$} This form captures multi-statement
code blocks, which occur in only 4\% of sites. These are larger expressions, we
do not inspect contents of the blocks. The median number of executed operations
is \packageMinimizedmedianoperationsjRnd. They typically run in function
environments.

@@

The environment in which evaluation happens determines what is visible to the
computation started by \eval and the potential reach of its side-effects. The
second argument to \eval specifies that environment. Environments of \eval can be
classified into the following four kinds:

\begin{compactitem}[---]
\item \emph{Function:} environments for the local variables of some function
  currently active on the call stack. Obtained by calling \c{parent.frame()} or
  \c{sys.frame()}.
\item \emph{Synthetic:} environments built from data structures such as lists,
  dataframes, or constructed explicitly with \c{new.env}, \c{list2env} or
  \c{as.environment}. Also includes the empty environment.
\item \emph{Global:}  environments in which scripts or interactive commands
  are evaluated.
\item \emph{Package:}  environments of a loaded library.
\end{compactitem}

\noindent
As shown in Table~\ref{tab:highlevelenvironments}, most calls evaluate in a
function environment with the \c{global} kind as a distant second. This means
that most variable lookups and most side effects either read/update existing
local variables or introduce new ones. But for which function? From the point of
view of the function that called \eval, Table~\ref{tab:funoffset} gives an
offset on the call stack. Thus 0 is the direct caller and corresponds to local variables; 1 is its parent, and so on. The data suggests that in 81\% of cases, \eval accesses its caller's environment -- this means, the variables of the function where \eval textually
occurs are read and written to. It is interesting that some 1.5\% of sites
evaluate a code in an environment that is three frames or above from \eval. This
distance implies that, in general, modular reasoning is impossible, to
understand the behavior of any piece of code entails fully understanding the
behavior of all functions that the code may call as the actions of \eval may
happen at a distance. Finally note that any particular site may have several
kinds of environments, but in \packageNbOneCategoryEnvirSitePercent of the case a
site has a single kind.

@@

\paragraph{Discussion}
The data presented in this section is what one could expect. ({\it Function})
The majority of \eval call sites access the environment of some function on the
call stack. Usually, the current function, but sometimes frames arbitrarily far
up the call stack. One thing the data does not say is how that environment was
obtained. The expected case is that the environment was the one obtained from a
promise; as promises combine code and the environment in which that code
originates. Likely less frequent are cases where \eval is provided the results
of programmatically selecting some call stack. ({\it Synthetic}) The relative
high frequency of synthetic environments corresponds to cases when one wants to
evaluate an expression in either a restricted environment, use a data
structure as an environment. ({\it Global}) The fact that the global environment
shows up so frequently is likely an artifact of the way the code is run. ({\it
  Package}) There are only 92 sites that use a package environment in \eval,
this is probably best as it is bad form to add bindings or modify functions
of a loaded package. Many more sites rather wrap a package environment in a new
environment to create singletons in a package.

@@

\begin{compactitem}[---]
\item {\it Reflection:} This group corresponds to uses of the \c{match.call}
  function to reflectively capture the expression that invoked the current
  function.
\item {\it Constructed:} Expressions can be constructed by invoking the
  \c{quote}, \c{enquote}, \c{expression}. Function arguments are passed as
  promises and \c{substitute} is used to retrieve the source expression
  associated with the promise.
\item {\it String:} Finally, expressions can be created from strings by invoking
   \c{parse}, \c{str2expression} or \c{str2lang}.
\end{compactitem}

\noindent
Table~\ref{tab:provenance} summarizes the expression provenance in our corpus.
This data is obtained by dynamically tainting values as they are produced by the
various sources, due to technical reasons, there is some imprecision in the
results. In particular, we are not able to classify all sites. Manual inspection
of numerous examples suggest that the classified results are accurate.

@@

\paragraph{Discussion}
The origin data suggests that constructing expressions from strings is a
minority of the use cases, instead the constructed category shows that the
majority of \evals comes from code that was processed by the compiler, and may be
slightly modified by the programmer before invoking \eval. Both constructed and
reflection categories roughly correspond to meta-programming. Some of these use
cases could likely be replaced by macros if the R designers could be convinced
to overcome their distaste for those.
